---
title: "PERMDISP"
output: html_document
date: "2025-02-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readxl)
library(nlme)
library(car)
library(vegan)
library(reshape2)
library(rstatix)
library(ggplot2)
library(gt)
knitr::opts_knit$set(root.dir = "C:/Users/mbkro/OneDrive/Documents/Krochta_JFE/GLS_PERMDISP")
```

**2024 SUBPERIOD 1**

**Setup and GLS**

*Daily mean*
```{r}
# Load dataset
sub24.1 <- read_excel("C:/Users/mbkro/OneDrive/Documents/Krochta_JFE/GLS_PERMDISP/Outlets_before8_1_2024.xlsx")

# Ensure date and site are properly formatted
sub24.1 <- sub24.1 %>%
  mutate(
    date = as.Date(date),
    site = as.factor(site)
  )

# Fit GLS model with AR(1) autocorrelation grouped by site
sub24.1_mn_model <- gls(
  daily_mean ~ Elevation + SiteAreaKm2 + tmean,
  data = sub24.1,
  correlation = corAR1(form = ~ date | site),
  method = "ML"
)

# View model summary
summary(sub24.1_mn_model)

# Extract normalized residuals and add to dataframe
sub24.1$MeanTemp_resid <- residuals(sub24.1_mn_model, type = "normalized")

# Plot autocorrelation of residuals
acf(sub24.1$MeanTemp_resid, main = "ACF of GLS Normalized Residuals")

# Check VIF using standard lm 
lm_temp <- lm(daily_mean ~ Elevation + SiteAreaKm2 + tmean, data = sub24.1)
vif(lm_temp)

# Visualize residuals
par(mfrow = c(2, 2))
plot(lm_temp)
```

*Daily max*
```{r}
# Fit GLS model with AR(1) autocorrelation grouped by site
sub24.1_mx_model <- gls(
  daily_max ~ Elevation + SiteAreaKm2 + tmean,
  data = sub24.1,
  correlation = corAR1(form = ~ date | site),
  method = "ML"
)

# View model summary
summary(sub24.1_mx_model)

# Extract normalized residuals and add to dataframe
sub24.1$MaxTemp_resid <- residuals(sub24.1_mx_model, type = "normalized")

# Plot autocorrelation of residuals
acf(sub24.1$MaxTemp_resid, main = "ACF of GLS Normalized Residuals")

# Check VIF using standard lm 
lm_temp <- lm(daily_max ~ Elevation + SiteAreaKm2 + tmean, data = sub24.1)
vif(lm_temp)

# Visualize residuals
par(mfrow = c(2, 2))
plot(lm_temp)  
```

*Daily range*
```{r}
# Fit GLS model with AR(1) autocorrelation grouped by site
sub24.1_rg_model <- gls(
  daily_range ~ Elevation + SiteAreaKm2 + tmean,
  data = sub24.1,
  correlation = corAR1(form = ~ date | site),
  method = "ML"
)

# View model summary
summary(sub24.1_rg_model)

# Extract normalized residuals and add to dataframe
sub24.1$DailyRange_resid <- residuals(sub24.1_rg_model, type = "normalized")

# Plot autocorrelation of residuals
acf(sub24.1$DailyRange_resid, main = "ACF of GLS Normalized Residuals")

# Check VIF using standard lm 
lm_temp <- lm(daily_range ~ Elevation + SiteAreaKm2 + tmean, data = sub24.1)
vif(lm_temp)

# Visualize residuals
par(mfrow = c(2, 2))
plot(lm_temp)  
```

**Run PERMDIST with residuals and visualize variability**
```{r}
# Create response matrix with residuals
response_resid_matrix <- as.matrix(sub24.1[, c("MeanTemp_resid", "MaxTemp_resid", "DailyRange_resid")])
response_resid_matrix <- scale(response_resid_matrix)  

# Compute Euclidean distance matrix
dist_matrix_resid <- vegdist(response_resid_matrix, method = "euclidean")

# Run PERMDISP to assess site-level temperature variability
permdisp_resid <- betadisper(dist_matrix_resid, sub24.1$site)

# Plot variability by site
boxplot(permdisp_resid, main="Site-Level Thermal Variability (Early Summer 2024)",
        ylab="Distance to Centroid", xlab="Site")
```

**Perform Permutation Test to determine if sites differ significantly**
```{r}
# Perform permutation test to see if sites differ significantly in variability
permdisp_test <- permutest(permdisp_resid, permutations = 999)

# Print results
print(permdisp_test)
```

**Build heatmap showing pairwise differences**
```{r}
# Ensure Site is a factor
df_dispersion <- data.frame(Site = as.factor(sub24.1$site), Dispersion = permdisp_resid$distances)

# Run pairwise Wilcoxon test with Bonferroni correction
pairwise_dispersion_test <- df_dispersion %>%
  pairwise_wilcox_test(Dispersion ~ Site, p.adjust.method = "bonferroni")

# Add significance stars
pairwise_dispersion_test <- pairwise_dispersion_test %>%
  mutate(Significance = case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    TRUE      ~ ""
  ))

# Convert to wide matrix format for heatmap
heatmap_matrix <- dcast(pairwise_dispersion_test, group1 ~ group2, value.var = "p")
rownames(heatmap_matrix) <- heatmap_matrix$group1
heatmap_matrix <- heatmap_matrix[, -1] 

# Fill NA values (diagonal) with 1
heatmap_matrix[is.na(heatmap_matrix)] <- 1

# Melt into long format for ggplot
heatmap_melted <- melt(as.matrix(heatmap_matrix), na.rm = TRUE)
colnames(heatmap_melted) <- c("Site1", "Site2", "p_value")

# Add significance stars to the melted data
heatmap_melted <- heatmap_melted %>%
  left_join(pairwise_dispersion_test, by = c("Site1" = "group1", "Site2" = "group2"))

ggplot(heatmap_melted, aes(x = Site1, y = Site2, fill = p_value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Significance), color = "black", size = 5) + 
  scale_fill_gradient(low = "red4", high = "white", limits = c(0, 1)) +
  labs(title = "Pairwise Site Variability Differences (Early Summer 2024)",
       x = "Site",
       y = "Site",
       fill = "p-value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*Pairwise plot for Sub1*
```{r}
# Prepare data
df_dispersion <- data.frame(Site = as.factor(sub24.1$site), Dispersion = permdisp_resid$distances)

pairwise_dispersion_test <- df_dispersion %>%
  pairwise_wilcox_test(Dispersion ~ Site, p.adjust.method = "bonferroni")

# Assign significance levels 
pairwise_dispersion_test <- pairwise_dispersion_test %>%
  mutate(significance_level = case_when(
    p < 0.001 ~ "< 0.001",
    p < 0.01  ~ "< 0.01",
    p < 0.05  ~ "< 0.05",
    TRUE      ~ NA_character_
  ))

# Create symmetric matrix of p-values
all_sites <- sort(unique(c(pairwise_dispersion_test$group1, pairwise_dispersion_test$group2)))

# Full grid for melting (to ensure upper triangle plotting)
grid <- expand.grid(Site1 = all_sites, Site2 = all_sites) %>%
  filter(as.character(Site1) < as.character(Site2))  # upper triangle only

# Join p-values and significance
heatmap_data <- grid %>%
  left_join(pairwise_dispersion_test, by = c("Site1" = "group1", "Site2" = "group2"))

# Plot
ggplot(heatmap_data, aes(x = Site1, y = Site2, fill = significance_level)) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c(
      "< 0.001" = "#990000",
      "< 0.01"  = "#cc4c4c",
      "< 0.05"  = "#e6b0b0"
    ),
    na.value = "white",  # Non-significant = blank
    drop = FALSE
  ) +
  labs(
    title = "Pairwise Site Variability Differences (Early Summer 2024)",
    x = "Site",
    y = "Site",
    fill = "p-value"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

```

*Pairwise table*
```{r}
# Create table
pairwise_dispersion_test %>%
  gt() %>%
  tab_header(title = "Early Summer 2024 Pairwise PERMDISP p-values",
             subtitle = "Lower values indicate significant differences") %>%
  fmt_number(columns = everything(), decimals = 3)
```

*Site dispersion table*
```{r}
# Extract PERMDISP site-level dispersion values
site_dispersion <- data.frame(Site = permdisp_resid$group,
                              Dispersion = permdisp_resid$distances)

# Summarize by site: mean and standard deviation
site_dispersion <- site_dispersion %>%
  group_by(Site) %>%
  summarise(
    Mean_Dispersion = mean(Dispersion, na.rm = TRUE),
    SD_Dispersion = sd(Dispersion, na.rm = TRUE)
  )

# Print the output to check
print(site_dispersion)
```

**2024 SUBPERIOD 2**

**Setup and GLS**

*Daily mean*
```{r}
# Load dataset
sub24.2 <- read_excel("C:/Users/mbkro/OneDrive/Documents/Krochta_JFE/GLS_PERMDISP/Outlets_after8_1_2024.xlsx")

# Ensure date and site are properly formatted
sub24.2 <- sub24.2 %>%
  mutate(
    date = as.Date(date),
    site = as.factor(site)
  )

# Fit GLS model with AR(1) autocorrelation grouped by site
sub24.2_mn_model <- gls(
  daily_mean ~ Elevation + SiteAreaKm2 + tmean,
  data = sub24.2,
  correlation = corAR1(form = ~ date | site),
  method = "ML"
)

# View model summary
summary(sub24.2_mn_model)

# Extract normalized residuals and add to dataframe
sub24.2$MeanTemp_resid <- residuals(sub24.2_mn_model, type = "normalized")

# Plot autocorrelation of residuals
acf(sub24.2$MeanTemp_resid, main = "ACF of GLS Normalized Residuals")

# Check VIF using standard lm 
lm_temp <- lm(daily_mean ~ Elevation + SiteAreaKm2 + tmean, data = sub24.2)
vif(lm_temp)

# Visualize residuals
par(mfrow = c(2, 2))
plot(lm_temp)  
```

*Daily max*
```{r}
# Fit GLS model with AR(1) autocorrelation grouped by site
sub24.2_mx_model <- gls(
  daily_max ~ Elevation + SiteAreaKm2 + tmean,
  data = sub24.2,
  correlation = corAR1(form = ~ date | site),
  method = "ML"
)

# View model summary
summary(sub24.2_mx_model)

# Extract normalized residuals and add to dataframe
sub24.2$MaxTemp_resid <- residuals(sub24.2_mx_model, type = "normalized")

# Plot autocorrelation of residuals
acf(sub24.2$MaxTemp_resid, main = "ACF of GLS Normalized Residuals")

# Check VIF using standard lm 
lm_temp <- lm(daily_max ~ Elevation + SiteAreaKm2 + tmean, data = sub24.2)
vif(lm_temp)

# Visualize residuals
par(mfrow = c(2, 2))
plot(lm_temp)  
```

*Daily range*
```{r}
# Fit GLS model with AR(1) autocorrelation grouped by site
sub24.2_rg_model <- gls(
  daily_range ~ Elevation + SiteAreaKm2 + tmean,
  data = sub24.2,
  correlation = corAR1(form = ~ date | site),
  method = "ML"
)

# View model summary
summary(sub24.2_rg_model)

# Extract normalized residuals and add to dataframe
sub24.2$DailyRange_resid <- residuals(sub24.2_rg_model, type = "normalized")

# Plot autocorrelation of residuals
acf(sub24.2$DailyRange_resid, main = "ACF of GLS Normalized Residuals")

# Check VIF using standard lm 
lm_temp <- lm(daily_range ~ Elevation + SiteAreaKm2 + tmean, data = sub24.2)
vif(lm_temp)

# Visualize residuals
par(mfrow = c(2, 2))
plot(lm_temp) 
```

**Creating model summary table**
```{r}
# Define manual R2 function with explicit data input
r_squared_gls_manual <- function(model, data, response_var) {
  fitted_vals <- fitted(model)
  observed_vals <- data[[response_var]]
  rss <- sum((observed_vals - fitted_vals)^2, na.rm = TRUE)
  tss <- sum((observed_vals - mean(observed_vals, na.rm = TRUE))^2, na.rm = TRUE)
  return(1 - (rss / tss))
}

# List 6 models with both data and response variables
gls_models_6 <- list(
  mean_early  = list(model = sub24.1_mn_model, data = sub24.1, response = "daily_mean"),
  max_early   = list(model = sub24.1_mx_model, data = sub24.1, response = "daily_max"),
  range_early = list(model = sub24.1_rg_model, data = sub24.1, response = "daily_range"),
  mean_late   = list(model = sub24.2_mn_model, data = sub24.2, response = "daily_mean"),
  max_late    = list(model = sub24.2_mx_model, data = sub24.2, response = "daily_max"),
  range_late  = list(model = sub24.2_rg_model, data = sub24.2, response = "daily_range")
)

# Summary extraction function
extract_gls_summary <- function(model_obj, model_name) {
  model <- model_obj$model
  data <- model_obj$data
  response_var <- model_obj$response
  coefs <- summary(model)$tTable
  coef_names <- rownames(coefs)

  # Identify temperature variable
  temp_var <- setdiff(coef_names, c("(Intercept)", "Elevation", "SiteAreaKm2"))[1]

  # Coefficient extractor
  safe_coef <- function(var, stat) {
    if (var %in% coef_names) coefs[var, stat] else NA
  }

  # Extract Phi (AR1 autocorrelation)
  phi <- tryCatch(
    coef(model$modelStruct$corStruct, unconstrained = FALSE)[1],
    error = function(e) NA
  )

  # Calculate R² manually using supplied data
  r2 <- r_squared_gls_manual(model, data, response_var)

  # Return row
  data.frame(
    Model = model_name,
    Intercept = safe_coef("(Intercept)", "Value"),
    Intercept_p = safe_coef("(Intercept)", "p-value"),
    Elevation = safe_coef("Elevation", "Value"),
    Elevation_p = safe_coef("Elevation", "p-value"),
    SiteAreaKm2 = safe_coef("SiteAreaKm2", "Value"),
    SiteAreaKm2_p = safe_coef("SiteAreaKm2", "p-value"),
    TempVar = temp_var,
    TempVar_Est = safe_coef(temp_var, "Value"),
    TempVar_p = safe_coef(temp_var, "p-value"),
    Phi = phi,
    Resid_SE = summary(model)$sigma,
    AIC = AIC(model),
    BIC = BIC(model),
    R2 = r2
  )
}

# Build summary table
model_summary_6 <- bind_rows(lapply(names(gls_models_6), function(name) {
  extract_gls_summary(gls_models_6[[name]], name)
}))

# Print and export final table
print(model_summary_6)
write.csv(model_summary_6, "GLS_model_summaries_with_r2.csv", row.names = FALSE)
```

**Run PERMDIST with residuals and visualize variablity**
```{r}
# Create response matrix with residuals
response_resid_matrix <- as.matrix(sub24.2[, c("MeanTemp_resid", "MaxTemp_resid", "DailyRange_resid")])
response_resid_matrix <- scale(response_resid_matrix)  

# Compute Euclidean distance matrix
dist_matrix_resid <- vegdist(response_resid_matrix, method = "euclidean")

# Run PERMDISP to assess site-level temperature variability
permdisp_resid <- betadisper(dist_matrix_resid, sub24.2$site)

# Plot variability by site
boxplot(permdisp_resid, main = "Site-Level Thermal Variability (Late Summer 2024)",
        ylab = "Distance to Centroid", xlab = "Site")

```

**Perform Permutation Test to determine if sites differ significantly**
```{r}
# Perform permutation test to see if sites differ significantly in variability
permdisp_test <- permutest(permdisp_resid, permutations = 999)

# Print results
print(permdisp_test)
```

**Build heatmap showing pairwise differences**
```{r}
# Ensure Site is a factor
df_dispersion <- data.frame(Site = as.factor(sub24.2$site), Dispersion = permdisp_resid$distances)

# Run pairwise Wilcoxon test with Bonferroni correction
pairwise_dispersion_test <- df_dispersion %>%
  pairwise_wilcox_test(Dispersion ~ Site, p.adjust.method = "bonferroni")

# Add significance stars
pairwise_dispersion_test <- pairwise_dispersion_test %>%
  mutate(Significance = case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    TRUE      ~ ""
  ))

# Convert to wide matrix format for heatmap
heatmap_matrix <- dcast(pairwise_dispersion_test, group1 ~ group2, value.var = "p")
rownames(heatmap_matrix) <- heatmap_matrix$group1
heatmap_matrix <- heatmap_matrix[, -1] # Remove group1 column

# Fill NA values (diagonal) with 1
heatmap_matrix[is.na(heatmap_matrix)] <- 1

# Melt into long format for ggplot
heatmap_melted <- melt(as.matrix(heatmap_matrix), na.rm = TRUE)
colnames(heatmap_melted) <- c("Site1", "Site2", "p_value")

# Add significance stars to the melted data
heatmap_melted <- heatmap_melted %>%
  left_join(pairwise_dispersion_test, by = c("Site1" = "group1", "Site2" = "group2"))

ggplot(heatmap_melted, aes(x = Site1, y = Site2, fill = p_value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Significance), color = "black", size = 5) + 
  scale_fill_gradient(low = "red4", high = "white", limits = c(0, 1)) +
  labs(title = "Pairwise Site Variability Differences (Late Summer 2024)",
       x = "Site",
       y = "Site",
       fill = "p-value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

*Pairwise plot (sub2)*
```{r}
# Prepare data
df_dispersion <- data.frame(Site = as.factor(sub24.2$site), Dispersion = permdisp_resid$distances)

pairwise_dispersion_test <- df_dispersion %>%
  pairwise_wilcox_test(Dispersion ~ Site, p.adjust.method = "bonferroni")

# Assign significance levels 
pairwise_dispersion_test <- pairwise_dispersion_test %>%
  mutate(significance_level = case_when(
    p < 0.001 ~ "< 0.001",
    p < 0.01  ~ "< 0.01",
    p < 0.05  ~ "< 0.05",
    TRUE      ~ NA_character_
  ))

# Create symmetric matrix of p-values
all_sites <- sort(unique(c(pairwise_dispersion_test$group1, pairwise_dispersion_test$group2)))

# Full grid for melting (to ensure upper triangle plotting)
grid <- expand.grid(Site1 = all_sites, Site2 = all_sites) %>%
  filter(as.character(Site1) < as.character(Site2))  # upper triangle only

# Join p-values and significance
heatmap_data <- grid %>%
  left_join(pairwise_dispersion_test, by = c("Site1" = "group1", "Site2" = "group2"))

# Plot
ggplot(heatmap_data, aes(x = Site1, y = Site2, fill = significance_level)) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c(
      "< 0.001" = "#990000",
      "< 0.01"  = "#cc4c4c",
      "< 0.05"  = "#e6b0b0"
    ),
    na.value = "white",  # Non-significant = blank
    drop = FALSE
  ) +
  labs(
    title = "Pairwise Site Variability Differences (Late Summer 2024)",
    x = "Site",
    y = "Site",
    fill = "p-value"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

```

*pairwise p-value table (sub2)*
```{r}
# Create a styled table
pairwise_dispersion_test %>%
  gt() %>%
  tab_header(title = "Late Summer 2024 Pairwise PERMDISP p-values",
             subtitle = "Lower values indicate significant differences") %>%
  fmt_number(columns = everything(), decimals = 3) 
```

*Site dispersion table*
```{r}
# Extract PERMDISP site-level dispersion values
site_dispersion <- data.frame(Site = permdisp_resid$group,
                              Dispersion = permdisp_resid$distances)

# Summarize by site: mean and standard deviation
site_dispersion <- site_dispersion %>%
  group_by(Site) %>%
  summarise(
    Mean_Dispersion = mean(Dispersion, na.rm = TRUE),
    SD_Dispersion = sd(Dispersion, na.rm = TRUE)
  )

# Print the output to check
print(site_dispersion)
```

**Combined boxplots**
```{r}
# Create dispersion data for Early Summer
dist_early <- vegdist(scale(as.matrix(sub24.1[, c("MeanTemp_resid", "MaxTemp_resid", "DailyRange_resid")])), method = "euclidean")
permdisp_early <- betadisper(dist_early, group = sub24.1$site)
site_disp_early <- data.frame(
  Site = sub24.1$site,
  Distance = permdisp_early$distances,
  Period = "Early"
)

# Create dispersion data for Late Summer
dist_late <- vegdist(scale(as.matrix(sub24.2[, c("MeanTemp_resid", "MaxTemp_resid", "DailyRange_resid")])), method = "euclidean")
permdisp_late <- betadisper(dist_late, group = sub24.2$site)
site_disp_late <- data.frame(
  Site = sub24.2$site,
  Distance = permdisp_late$distances,
  Period = "Late"
)

# Combine both into one dataframe
dispersion_combined <- rbind(site_disp_early, site_disp_late)

# Plot 
p <- ggplot(dispersion_combined, aes(x = Site, y = Distance, fill = Period)) +
  geom_boxplot(position = position_dodge(width = 0.8), linewidth = 0.4) +  
  labs(x = "", y = "Dispersion") +
  scale_fill_manual(values = c("Early" = "#73bfe6", "Late" = "#33a02c")) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 16),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.key.size = unit(0.6, "cm"),      # smaller legend boxes
    legend.key.width = unit(0.6, "cm"),
    legend.key.height = unit(0.6, "cm"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )

ggsave("dispersion_plot.png", plot = p,
       width = 8, height = 6, dpi = 600, bg = "white")


# High-resolution PNG with white background
ggsave("dispersion_plot.png", plot = p,
       width = 8, height = 6, dpi = 600, bg = "white")
```

